---
title: "Data Manipulation and cleaning"
author: "Divy Joshi"
output: word_document
---

## Two ways to prepare Dataset

#1. Type the variables and the data in 

```{r}
#1. type the variables and the data in

#define the variables
x=1:10
y=c(2,3,5,8,10,12,13,16,18,20)

#create a data frame
data=data.frame(x,y)

#check the dataset
View(data)

#check the variables in the data
names(data)

#scatterplot
plot(x,y)

#fit a simple linear regression model
model1=lm(y~x)

#add the fitted model line into the scatterplot
abline(model1)

```

# BASICS - Data Manipulation

```{r}
#data()

library(readxl)
BMI <- read_excel("C:/Users/divyj/Downloads/BMI.xlsx")
View(BMI)
head(BMI)
```

```{r}
# Tidyverse Libraray

library(tidyverse)

colnames(BMI)[4] = "Weight"
colnames(BMI)[3] = "Height"


BMI %>%  # %>% to continue the code
  select(Gender, Weight, Height , Age) %>%
  filter(Age > 20) %>%
  na.omit() %>% # To remove the null values
  mutate(Height = replace(Height, Height == 5.8, 176)) %>% # Replacing a Value
  mutate(Height = Height/100) %>%
  mutate(BMI_Calc = Weight/Height^2) %>%
  group_by(Gender) %>%
  summarise(Average_BMI = mean(BMI_Calc))

BMI$Height[BMI$Height == 5.8] <- 176
BMI
```

```{r}

# Adding a New Column

descr <- c()

for (i in BMI$Height) {
if (i < 155) {
  descr <- append(descr,"Small")
} else if (i > 155 & i < 165) {
  descr <- append(descr,"Meduim")
} else {
  descr <- append(descr,"Large")
}
}
descr

new_col = data.frame(descr)
new_col

# Adding Column to Final data

BMI_F <- cbind(BMI,new_col)
BMI_F
```

# Types of DAta

```{r}
str(BMI_F)
```

```{r}

# Changing the data type

BMI_F$descr <- as.factor(BMI_F$descr)
BMI_F$Age <- as.integer(BMI_F$Age)
levels(BMI_F$descr)
str(BMI_F)
```

```{r}

# Changing the Levels

BMI_F$descr <- factor(BMI_F$descr, 
                      levels = c("Small","Meduim","Large"))
levels(BMI_F$descr)
```

```{r}

# Work on Vectors

# BMI_F$Age # is a vector
BMI_F$Old <- BMI_F$Age > 23 #Creating a new column
BMI_F
class(BMI_F$Old)
str(BMI_F)
```

# Renaming Variables and Recorder Columns

```{r}

new <- BMI_F %>%
  select(Age, Height, Weight, Gender) %>%
  rename(mass = Weight)
new
BMI_F
```

# Recoding Data - Without using If else

```{r}
library(dplyr)

sw <- BMI_F %>%
  select(Age, Height, Weight, Gender, Old) %>%
  na.omit() %>%
  mutate(Height = Height / 100) %>%
  #filter(Gender == "Male" | 
           #Gender == "Female") 
  filter(Gender %in% c("Male", "Female")) %>% # Can filter more than one
  mutate(Gender = recode(Gender,Male = "M", Female = "F")) %>%
  mutate(size = Height > 1 & Weight > 75,
         size = if_else(size == TRUE, "big", "small"))
sw       
```

# Filter

```{r}
data("msleep")
View(msleep)
```

```{r}

my_data <- msleep %>%
  select(name, sleep_total, order) %>%
#  filter(sleep_total > 18)
  filter(!sleep_total > 18)
my_data
```

```{r}

my_data1 <- msleep %>%
  select(name,order,bodywt, sleep_total) %>%  
# filter(order == "Primates", bodywt > 20)   #and
# filter(order == "Primates" | bodywt > 20)  #OR
  filter(order == "Primates" & bodywt > 20) #and
my_data1
```

```{r}

my_data2 <- msleep %>%
  select(name, sleep_total) %>%
  filter(name == "Cow" |
           name == "Dog" |
           name == "Goat")
my_data2

```

```{r}

#Concatenation

my_data3 <- msleep %>%
  select(name, sleep_total) %>%
  filter(name %in% c("Cow", "Dog", "Horse"))
my_data3
```

```{r}

# Between

my_data4 <- msleep %>%
  select(name, sleep_total) %>%
  filter(between(sleep_total, 16, 18))
my_data4
```

```{r}

# Near - How close

library(tidyverse)
my_data5 <- msleep %>%
  select(name, sleep_total) %>%
  filter(near(sleep_total, 17, tol = 0.5))
my_data5
```

```{r}

# Null values

my_data6 <- msleep %>%
  select(name, conservation, sleep_total) %>%
  filter(!is.na(conservation))
my_data6
```

# Exploring Data

```{r}

dim(msleep) # Rows and Columns
str(msleep) # Data types
glimpse(msleep)
head(msleep)
tail(msleep)

```

```{r}

attach(msleep)
names(msleep)
class(genus)
length(sleep_total)
unique(conservation)
```

```{r}
View(sort(table(conservation), decreasing = TRUE))
barplot(sort(table(conservation), decreasing = TRUE))
```

```{r}
# Same above in tidyverse

msleep %>%
  select(conservation) %>%
  count(conservation) %>%
  arrange(desc(n)) %>%
  View()
```

# Missing Values

```{r}

#msleep[,5]

View(msleep[is.na(conservation), ])
```

# Data Manipulation

```{r}
# Renaming  a varibale

msleep %>% 
  rename("conserv" = "conservation") %>%  # New Name - Existing Name
  glimpse()

# Re order Variable

msleep %>% 
  select(vore, name, everything())

# Change a varible type

msleep$vore <- as.factor(msleep$vore)
glimpse(msleep)

## --  Same with tidyverse -- 

msleep %>% 
  mutate(vore = as.character(vore)) %>% 
  glimpse()

# Select varaibles to work with

names(msleep)

msleep %>% 
  select(2:4,
         awake,
         starts_with("sleep"),
         contains("wt")) %>% 
  names()


# Filter and Arranging 

unique(msleep$order)

msleep %>% 
  filter((order == "Carnivora" |
          order == "Primates" ) & sleep_total > 8 ) %>% 
  select(name, order, sleep_total) %>% 
  arrange(-sleep_total) %>% 
  View

### --- Another Way ---

msleep %>% 
  filter(order %in% c("Carnivora", "Primates") & sleep_total >  8) %>% 
  select(name, order, sleep_total) %>% 
  arrange(order) %>% 
  View

# Change Observations 

msleep %>% 
  mutate(brainw_in_grams = brainwt * 1000) %>% 
  View

# COnditional Change using If Else

size_brain <- msleep %>% 
  select(name, brainwt) %>% 
  drop_na(brainwt) %>% 
  mutate(brain_size = if_else(brainwt > 0.01, "Large", "Small"))
size_brain


# Recoding Data

size_brain %>% 
  mutate(size_brain = recode(brain_size, "Large" = 1, "Small" = 2))
size_brain
```

# Reshaping Data from wide to long or long to wide\_

```{r}

library(gapminder)
View(gapminder)

data <- select(gapminder, country, year, lifeExp)
data

```

```{r}

# Wide data - Transposing Data

wide_data <- data %>% 
  pivot_wider(names_from = year, values_from = lifeExp)

View(wide_data)

long_data <- wide_data %>% 
  pivot_longer(2:13,
               names_to = "year",
               values_to = "lifeExp")
View(long_data)
```

# Describing and Summarising Data

```{r}
library(tidyverse)

glimpse(msleep)

min(msleep$awake)
max(msleep$awake)
range(msleep$awake)
IQR(msleep$awake)
mean(msleep$awake)
median(msleep$awake)
var(msleep$awake)

```

```{r}
summary(msleep)
summary(msleep$sleep_total)
```

```{r}

msleep %>% 
  select(sleep_total, brainwt) %>% 
  summary
```

```{r}

# Creating a summary table

msleep %>% 
  drop_na(vore) %>% 
  group_by(vore) %>% 
  summarise(Lower = min(sleep_total),
            Average = mean(sleep_total),
            Upper = max(sleep_total),
            Difference = max(sleep_total) - min(sleep_total)) %>% 
  arrange(Average) %>% 
  View()

```

```{r}

# Creating Contingency table

library(MASS)
attach(Cars93)
glimpse(Cars93)
```

```{r}

table(Origin)
table(AirBags, Origin)
addmargins(table(AirBags, Origin),1) # 1- rows, 2 - columns
```

```{r}
prop.table(table(AirBags,Origin))*100 # Try 1 and 2 after ", " tht add up 100 %
```

```{r}
round(prop.table(table(AirBags, Origin), 2)*100)
```

```{r}

# -- All of above in tidyverse --

Cars93 %>% 
  group_by(Origin, AirBags) %>% 
  summarise(number = n()) %>% # Counting how many
  pivot_wider(names_from = Origin, values_from = number)
```

# working with Factors and categorial Variables - Package [ Forcast ]

```{r}

library(tidyverse)
library(forcats)
library(patchwork)

?forcats

View(gss_cat) # Inbuilt data in Forcats
```

```{r}
glimpse(gss_cat)
```

```{r}
unique(gss_cat$race)
```

```{r}
gss_cat %>% 
  pull(race) %>%  #Pulling race variable out as a vector, as do not want to see it as a dataframe
  unique()
```

```{r}
# Counting
count(gss_cat, race, sort = T) # dataframe, vector, sort
```

```{r}
# Same above

gss_cat %>% 
  count(race)

# Levels

gss_cat %>% 
  pull(race) %>% 
  levels()

# Table

gss_cat %>% 
  pull(race) %>% 
  table()
```

```{r}

# If we want to get rid of factor - "Not applicable" thats not usable at all

# Using package Forcat

gss_cat %>% 
  mutate(race = fct_drop(race)) %>% 
  pull(race) %>% 
  levels()
```

```{r}
# Order a factor level by value
# of another numeric value

gss_cat %>% 
  drop_na(tvhours) %>% 
  group_by(relig) %>%
  summarise(mean_tv = mean(tvhours)) %>% 
  ggplot(aes(mean_tv, relig)) +
  geom_point(size = 4)
```

```{r}

# Exact same code above, only reordering religion by fct_order [ Forcart ]

gss_cat %>% 
  drop_na(tvhours) %>% 
  group_by(relig) %>% 
  summarise(mean_tv = mean(tvhours)) %>% 
  mutate(relig = fct_reorder(relig, mean_tv)) %>% 
  ggplot(aes(mean_tv, relig)) +
  geom_point(size = 5)
```

```{r}

# Reverse Factor Levels - Income graph

gss_cat %>% 
  drop_na(age) %>% 
  filter(rincome != "Not applicable") %>% 
  group_by(rincome) %>% 
  summarise(mean_age = mean(age)) %>% 
  ggplot(aes(mean_age, rincome)) +
  geom_point(size = 4)
```

```{r}

# Reordering above code

gss_cat %>% 
  drop_na(age) %>% 
  filter(rincome != "Not applicable") %>% 
  group_by(rincome) %>% 
  summarise(mean_age = mean(age)) %>% 
  mutate(rincome = fct_rev(rincome)) %>% 
  ggplot(aes(mean_age, rincome)) +
  geom_point(size = 4)
```

```{r}

# order factor levels by frequency of the value of that variable

gss_cat %>% 
  count(marital)
```

```{r}

gss_cat %>% 
  ggplot(aes(marital)) + geom_bar()

```

```{r}

gss_cat %>% 
  mutate(marital = fct_infreq(marital)) %>% # Order frequency
  mutate(marital = fct_rev(marital)) %>% # Order frequency
  count(marital)
```

```{r}

gss_cat %>% 
  mutate(marital = marital %>% fct_infreq() %>% fct_rev()) %>% 
  ggplot(aes(marital))+
  geom_bar(fill = "steelblue", aplha = 0.5) + 
  theme_bw()
```

```{r}

unique(gss_cat$partyid)

# Recode Factors

gss_cat %>% 
  mutate(partyid = fct_recode(partyid,
                               "Other" = "No answer",
                               "Other" = "Don't know",
                               "Other" = "Other party")) %>% 
  count(partyid)
```

```{r}

# Factor Collapse - dose the same thing as above

gss_cat %>% 
  mutate(partyid = fct_collapse(partyid,
                                other = c("No answer", "Don't know", "Other party"))) %>% 
  count(partyid)
```

```{r}

# Lumping into Other - need only first two then all goes in Other

gss_cat %>% 
  count(relig, sort = T)

gss_cat %>% 
  mutate(relig = fct_lump(relig, n = 2)) %>% 
  count(relig)

```

```{r}

# Reordering a factor y by its value corresponding to the largest x value

gss_cat %>% 
  filter(!is.na(age)) %>% 
  filter(marital %in% c("Never married",
                        "Married",
                        "Widowed")) %>% 
  count(age, marital) %>% 
  group_by(age) %>% 
  mutate(prop = n / sum(n)) %>% 
  ggplot(aes(age, prop, colour = marital))+
  geom_line(sie = 2, na.rm = TRUE)+
  theme_minimal()
  
```

```{r}
boxplot(gss_cat$marital)
```

# Group By and Summarise

```{r}
names(msleep)
```

```{r}
msleep %>% 
  drop_na(sleep_rem, vore) %>% 
  group_by(vore) %>% 
  summarise('Average total sleep' = mean(sleep_total),
            'Maximum rem sleep' = max(sleep_rem)) %>% 
  head
```

# Seperate and Unite data

```{r}
library(gapminder)
View(gapminder)

# Seperate year into century and year

gapminder1 <- gapminder %>% 
  separate(col = year,
           into = c("century", "year"),
           sep = 2)
head(gapminder1)
```

```{r}

# Combining centuary and year into date

gapminder1 %>% 
  unite(col = date,
        century, year,
        sep = "")
```

# Joins

```{r}

library(dplyr)
data1 <- data.frame(ID = 1:2,                      # Create first example data frame
                    X1 = c("a1", "a2"),
                    stringsAsFactors = FALSE)
data2 <- data.frame(ID = 2:3,                      # Create second example data frame
                    X2 = c("b1", "b2"),
                    stringsAsFactors = FALSE)

inner_join(data1, data2, by = "ID")
left_join(data1, data2, by = "ID")
right_join(data1, data2, by = "ID")
full_join(data1, data2, by = "ID")
semi_join(data1, data2, by = "ID")
anti_join(data1, data2, by = "ID")


```
